---
title: "Parks in NYC"
author: "Connor Wolosewicz"
date: "4/2/26"
format:
  html:
    theme: cerulean
    toc: true
    toc-location: right
    toc-title: "Table of Contents"
    code-fold: false
    smooth-scroll: true
---

> *Environmental Justice in NYC: There's more than just Central Park.*


```{r}
#| message: false
#| warning: false
#| echo: false
library(tidyverse)
parks <- read_csv("Parks_Properties_20260320.csv") 
```

# Understanding New York City's Parks

For this data analysis, I chose to investigate data on the New York City Park system. This covers parks that are managed partially or solely by NYC Parks, including what neighborhood and zip code the park belongs to, what kind of buildings you can find there, and how close it is to the waterfront. Each row is a property, with NYC Parks owning over 2,000 as of this year. You can find this dataset [here.](https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve/about_data)

I am interested in a birds-eye view of NYC's park system, particularly to discover if parks are evenly distributed across the city's neighborhoods. As someone who hopes to go into urban planning and making cities more sustainable in the face of climate change, this data is a great window into understanding NYC's climate resilience and environmental justice. 

## Variables

I plan to hone in on several available variables, seen below.

| Variable | Type | Description |
|---|---|---|
| `ACRES` | Continuous | Size of each park property in acres |
| `BOROUGH` | Categorical | NYC borough where the park is located (B = Brooklyn, M = Manhattan, Q = Queens, R = Staten Island, X = Bronx) |
| `TYPECATEGORY` | Categorical | Type of park (e.g., Neighborhood Park, Triangle/Plaza, Nature Area) |

## Data Cleaning and Statistics

```{r}
#| echo: false
parks_clean <- parks |>
  select(ACRES, BOROUGH, TYPECATEGORY, WATERFRONT) |>
  mutate(
    BOROUGH = recode(BOROUGH,
      "B" = "Brooklyn",
      "M" = "Manhattan",
      "Q" = "Queens",
      "R" = "Staten Island",
      "X" = "Bronx"),
    BOROUGH = as.factor(BOROUGH),
    TYPECATEGORY = as.factor(TYPECATEGORY)
  ) |>
  filter(!is.na(ACRES)) |>
  group_by(TYPECATEGORY) |>
  filter(n() >=10) |>
  ungroup()
```

For this dataset, I dropped all of the other variables, and filtered out rows where the **ACRES** value was missing. Also, I made sure that the **BOROUGH** values were spelled out correctly so that they are easier to interpret. Finally, my parks_clean dataset additionally changed **BOROUGH** and **TYPECATEGORY**, categorical variables, to factors, ensuring they are treated as such and not characters. This ultimately makes the data easier to analyze. 

## **ACRES** Visualization

```{r}
#| include: false
summary(parks_clean$ACRES)
IQR(parks_clean$ACRES, na.rm = TRUE)

q1 <- quantile(parks_clean$ACRES, 0.25, na.rm = TRUE)
q3 <- quantile(parks_clean$ACRES, 0.75, na.rm = TRUE)
IQR_val <- IQR(parks_clean$ACRES, na.rm = TRUE)

upper <- q3 + 1.5 * IQR_val
lower <- q1 - 1.5 * IQR_val

sum(parks_clean$ACRES > upper|parks_clean$ACRES < lower, na.rm = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
parks_clean |>
  ggplot(aes(x = ACRES)) +
  geom_histogram(binwidth = 0.5, fill = "darkgreen", color = "white") +
  xlim(0,15) + 
  geom_vline(xintercept = median(parks_clean$ACRES, na.rm = TRUE), 
            color = "red", linetype = "dashed", linewidth=1) +
  labs(
    title = "Distribution of NYC Park Sizes",
    x = "Park Size (Acres)",
    y = "Number of Parks",
    caption = "Source: NYC Open Data - Parks Properties") + theme_minimal()
```

The distribution of park size in NYC is strongly right-skewed and unimodal, with the vast majority of parks being under 2 acres, and the largest park being 2,772 acres. In regards to other measures, the median is 0.70 acres and the IQR is 1.89 acres, demonstrating that the typical NYC park is quite small. Outside of this range lies only 310 outliers, underscoring the trend of small parks; 179 were removed from the histogram for legibility.

## **BOROUGH** and **TYPECATEGORY** Visualizations

```{r}
#| message: false
#| warning: false
#| echo: false
parks_clean |>
  ggplot(aes(x = BOROUGH)) +
  geom_bar(fill = "darkgreen", color = "white") +
  labs(title = "Number of Park Properties by Borough",
    x = "Borough",
    y = "Number of Parks",
    caption = "Source: NYC Open Data - Parks Properties"
  ) + theme_minimal()
```

```{r}
#| include: false
parks_clean |>
  count(BOROUGH) |>
  arrange(desc(n))
```

```{r}
#| message: false
#| warning: false
#| echo: false
parks_clean |>
  ggplot(aes(x = TYPECATEGORY)) +
  geom_bar(fill = "darkgreen", color = "white") +
  coord_flip() +
  labs(title = "Number of Park Properties by Type",
    x = "Park Type",
    y = "Number of Parks",
    caption = "Source: NYC Open Data - Parks Properties"
  ) + theme_minimal()
```

The above bar charts illustrate a complex park system. The number of parks are distributed across the city's boroughs fairly evenly, but there is a large gap between the greenest neighborhood (Brooklyn) and the least green (Staten Island). Specifically, Brooklyn has 463 more parks than Staten Island. Perhaps uniquely, gardens, playgrounds, neighborhood parks, and triangles/plazas are the most common types of public parks in NYC. Gardens in public parks is the most surprising to me, as I don't see that often in DC, but I see plenty of triangles caused by our diagonal avenues.  

## Comparing Variables—Boroughs by Park Size

```{r}
#| message: false
#| warning: false
#| echo: false
parks_clean |>
  ggplot(aes(x = BOROUGH, y = ACRES)) +
  geom_boxplot(fill = "darkgreen", color = "black") +
  ylim(0, 15) +
  labs(
    title = "Park Size by Borough",
    x = "Borough",
    y = "Park Size (Acres)",
    caption = "Source: NYC Open Data - Parks Properties"
  ) +
  theme_minimal()
```

To understand my initial query on how park size varies by borough, I compared these two variables in a boxplot. Interestingly, while having the least amount of parks, Staten Island stands out with both the highest median park size and the most spread. On the other hand, the Bronx, Brooklyn, and Manhattan all have very similar and small median park sizes, suggesting that most NYC Parks properties in these boroughs are tiny pocket parks. Taken together, all five boroughs show lots of outliers above the whiskers, showing that a small number of very large parks are present across the city but are far from typical. Overall, this pattern makes sense given NYC's density, as Staten Island is the least urbanized and has more room for larger green spaces.

*Note—I again removed 179 entries for legibility.*


